---
title: "Lab QA Report"
author: "Water Monitoring and Assessment"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

<!--
THANK YOU, MARY.

HAVE A FANTASTIC SUMMER.

BEST,
JIMMY
--> 

<!--
This markdown file performs a mass of QA operations and then generates a technical report.
The report will only have flagged ids show up.
--> 
```{r setup, echo=FALSE}
library('knitr')
library('DT')
#display labID
paste('Lab Accession Sequence Number:', lab_sequence)
```

# Introduction
The goal of this technical report is to run several scripts on a specific lab accession sequence number. The purpose of these scripts is to automate quality assurance tests to assist CT-DEP staff in dertermining if the information is consistent with the data quality objectives of their projects.

The folllowing tests will be run in this report:

* Count N/A values
* Flag Values compared to Milli-Q
* Flag duplicate values compared to lab precision rate
* Values entered as less than the minimum detection limit and not N/A
* 'Total' values against the sum of their component chemicals
* Values that are above the 95th percentile by basin
* Values that are above the 95th percentile by impervious cover grouping

## Languages and Software Used
The languages used were SQLite and R. The main software and IDEs used were SQLiteStudio, RStudio, QGIS, and ArcGIS. The R libraries used, in order of frequency, are RSQLite, ggplot2, and plyr.

## Notes & Data Manipulation
An important note is that many of the following scripts convert null values to the chemical parameter's defined minimum detection limit value. Another important note is that most of the scripts rely on SQL queries that stress taking only 'River/Stream' station types and no marked duplicate records. All lab accession number IDs are also shortened to the first six digits. Finally, all generated tables are in CSV (Comma-Separated Values) format.


<!-- Lab Accession Sequence(s)
This code block is where the user defines the desired lab accession ID(s) for this technical report.

The block will convert all tables to hold records of only the desired ID(s). -->
```{r takeDesiredLabAccessions, include=FALSE}
#set all appropiate data tables to just the chosen lab accessions  

table_chem <-  table_chem_base[table_chem_base$lab_accession %in% labIDs[z], ]
table_milliQ <-  table_milliQ_base[table_milliQ_base$lab_accession %in% labIDs[z], ]
table_sumOfParts <-  table_sumOfParts_base[table_sumOfParts_base$lab_accession %in% labIDs[z], ]
table_pairs <-  table_pairs_base[table_pairs_base$lab_accession %in% labIDs[z], ]
table_lessThanMDL <-  table_lessThanMDL_base[table_lessThanMDL_base$lab_accession %in% labIDs[z], ]

#set v_chems
#take only chemical parameters within this lab ID
v_chems <- c(unique(table_chem$chemparameter))
```

# The Database
The major tables in the database are as follows:

* chemdata
* sites
* basin
* table_QA
* mdl

## Chemical Data Table
The chemdata table holds the values of all the parameters.

```{r chem-table, echo=FALSE}
datatable(head(table_chem_base), rownames = FALSE, options = list)
```

* sta_seq: The station sequence number
* lab_accession: The lab accession sequence number
* collect_date: The date the sample was collected
* chemparameter: The chemical parameter
* value:
* uom: The unit of measurement
* station_type: The station type, this report uses only 'River/Stream' data
* duplicate: A binary value that defines if the sample is a testing duplicate (1) or not (0)

## Sites Table
The sites table holds site information.

```{r sites-table, echo=FALSE}
datatable(head(sitebasin_base), rownames = FALSE, options = list)
```

* name: The name of the stream
* ylat: The latitude
* xlong: The longitude
* sbasn: The sub-basin
* major mbasn: The name of the major basin

## Basin Table
The basin table 

```{r basin-table, echo=FALSE}
datatable(head(table_basin_base), rownames = FALSE, options = list)
```

## QA Table
This is a table of current values in the QA table within the database. The chemical parameters listed here are the only ones tested in sections 3.2 and 3.3.

```{r QA-Table, echo=FALSE}
#take only chemical parameters in table_milliQ
milli_chem <- c(unique(table_milliQ$chemparameter))
#check for chemical parameters in table_milliQ compared to table_QA
#if there is no corresponding parameter in table_QA, toss
milli_chem <- intersect(milli_chem, unique(table_QA$chemparameter))

#display
datatable(table_QA, rownames = FALSE, options = list)
```

* field_duplicate_precision: The field precision rate to compare duplicate values
* lab_duplicate_precision: The lab precision rate to compare Milli-Q values

## MDL Table
The MDL values were obtained page 9 of the Quality Assurance Project Plan (QAPP) Aquatic Life Response to Cultural Eutrophication in CT Freshwater Rivers and Streams (2017 - 2020).

```{r mdl-table, echo=FALSE}
datatable(head(table_mdl), rownames = FALSE, options = list)
```

* MDL: The minimum detection limit

# Script Results

## Count N/A Values
The following is a table that contains the count of null values by chemical parameter.

```{r countNull, include=FALSE}
#Join in R syntax to create a table with sites and chemicals
table_basin<-merge(x=table_chem, y=sitebasin_base, by="sta_seq", all.x=TRUE)
#declare and wipe data frame null clean for resuse
df_null <- NULL
#loop until end of amound of chemical parameters
for (n in 1:length(v_chems)) {
  chem = v_chems[n]
  ##Parse out some data by row
  #df of all values
  n_total <-table_basin[table_basin$chemparameter==chem & table_basin$duplicate==0, ]
  #df of null values
  n_NA <-table_basin[table_basin$chemparameter==chem & table_basin$duplicate==0 & table_basin$value =='NA', ]
  
  #create list of values omiting NA
  l_isNumeric <- na.omit(n_total$value)
  l_isNumeric <- sapply(l_isNumeric, '[', seq(max(sapply(l_isNumeric, length))))
  
  df_nullTempRow <- data.frame(chemical = chem,
                               total = c(nrow(n_total)),
                               countNumeric = c(length(l_isNumeric)),
                               countNA = c(nrow(n_NA)))
  
  df_null <- rbind(df_null, df_nullTempRow)
}
```
```{r tables-df_null, fig.cap='Table 1. Count of N/A values by chemical', fig.align="center", echo=FALSE}
datatable(df_null, rownames = FALSE, options = list)
```


## Flagged Milli-Q Values
The following is a table of flagged Milli-Q values where the percent difference between it and the minimum detection limit was the lab precision rate.

```{r flaggedMilli, include=FALSE}
df_total <- NULL #reset df_total at runtime

#loop until end of amound of chemical parameters
for (n in 1:length(milli_chem)) {
  #Set chosen chemical
  chemChoice <- milli_chem[n]
  #create chem dataframe of specific chemical parameter
  chem <- table_milliQ[which(table_milliQ$chemparameter == chemChoice), ]
  
  if (nrow(chem) > 0) { #check that chem has records
    #get lab duplicate precision rate for chemparameter
    RPD <- table_QA$lab_duplicate_precision[table_QA$chemparameter == chemChoice]
    
    #convert null milliQ values to mdl
    mdl <- chem$MDL[1]
    chem$Milli_Q[is.na(chem$Milli_Q)] <- mdl
    
    #add columns/fields to mark if pair failed RPD test
    chem <- cbind(chem, 'lab_precision_rate' = c(RPD))
    chem <- cbind(chem, "Percent_Diff" = c(0.0))
    
    #percent difference
    chem$Percent_Diff <- c(
      abs(chem$Milli_Q - chem$MDL) /
        ((chem$Milli_Q + chem$MDL) / 2)
    )
    
    #TRUE if Percent_Diff is greater than RPD (15%)
    #keep only records where Percent_Diff is higher
    chem <- chem[(chem$Percent_Diff > chem$lab_precision_rate), ]
    
    #add chem df to total dataframe
    df_total <- rbind(df_total, chem)
  }
  
}

#-----END-LOOP------#
if (!is.null(nrow(df_total)) == TRUE && nrow(df_total) > 0) {
  #need to convert $collect_date to date format
  df_total$collect_date <- as.Date(df_total$collect_date, "%m/%d/%Y")
  # sort
  df_total <- df_total[with(df_total, order(chemparameter, collect_date, sta_seq)), ]
  #convert to percent format for viewing
  df_total$Percent_Diff <- signif(df_total$Percent_Diff, digits = 4) #round
  df_total$lab_precision_rate <- signif(df_total$lab_precision_rate, digits = 4)
  df_total$Percent_Diff <- paste(df_total$Percent_Diff * 100, "%")
  df_total$lab_precision_rate <- paste(df_total$lab_precision_rate * 100, "%")
}
```
```{r tables-df_milliQ, fig.cap='Table 2. Count of flagged Milli-Q values', fig.align="center", echo=FALSE}
datatable(df_total, rownames = FALSE, options = list(pageLength = 5))
```


## Flagged Duplicate Values
The following is a table of flagged duplicate values where the differences between them was above the field precision rate.

```{r flaggedDup, include=FALSE}
df_total <- NULL #reset df_total at runtime

#loop until end of amound of chemical parameters
for (n in 1:length(milli_chem)) {
  #Set chosen chemical
  chemChoice <- milli_chem[n]
  #create chem dataframe of specific chemical parameter
  chem <- table_pairs[which(table_pairs$chemparameter == chemChoice), ]
  
  if (nrow(chem) > 0) { #check if chem has any values
    #Create MDL table
    test_table_mdl <- table_mdl[table_mdl$chemparameter == chemChoice & table_mdl$MDL, ]
    mdl <- test_table_mdl[1, 2] # convert value to MDL numeric vector
      
    #get FIELD duplicate precision rate for chemparameter
    RPD <- table_QA$field_duplicate_precision[table_QA$chemparameter == chemChoice]
    
    #convert all NULL or N/A values to chemical's MDL (minimum detection limit)
    chem$value_field[is.na(chem$value_field)] <- mdl
    chem$value_duplicate[is.na(chem$value_duplicate)] <- mdl
    
    #add columns
    chem <- cbind(chem, 'field_precision_rate' = c(RPD))
    chem <- cbind(chem, "Percent_Diff" = c(0.0))
    
    #percent difference
    chem$Percent_Diff <- c(
      abs(chem$value_field - chem$value_duplicate) /
        ((chem$value_field + chem$value_duplicate) / 2))
    
    #keep records where percent difference above field precision rate
    chem <- chem[(chem$Percent_Diff > chem$field_precision_rate), ]
    
    #add chem df to total dataframe
    df_total <- rbind(df_total, chem)
  }
  
}
#-----END-LOOP------#
if (!is.null(nrow(df_total)) == TRUE && nrow(df_total) > 0) {
  #need to convert $collect_date to date format
  df_total$collect_date <- as.Date(df_total$collect_date, "%m/%d/%Y")
  # sort
  df_total <- df_total[with(df_total, order(chemparameter, collect_date, sta_seq)), ]
  #convert to percent format for viewing
  df_total$Percent_Diff <- signif(df_total$Percent_Diff, digits = 4) #round
  df_total$field_precision_rate <- signif(df_total$field_precision_rate, digits = 4)
  df_total$Percent_Diff <- paste(df_total$Percent_Diff * 100, "%")
  df_total$field_precision_rate <- paste(df_total$field_precision_rate * 100, "%")
}
```
```{r tables-df_duplicate, fig.cap='Table 3. Count of flagged duplicate values', fig.align="center", echo=FALSE}
datatable(df_total, rownames = FALSE, options = list(pageLength = 5))
```


## Values Less Than MDL and Not NULL
The following is a table of values less than the minimum detection limit and are not NULL.

```{r lessThanMDL, include=FALSE}
#copy only rows where value is less than MDL
lessThanMDL_final <- table_lessThanMDL[which(table_lessThanMDL$value < table_lessThanMDL$MDL), ]
#omit null rows
lessThanMDL_final <- na.omit(lessThanMDL_final)
#shorten to specific fields if there is data
if (!is.null(nrow(lessThanMDL_final)) == TRUE && nrow(lessThanMDL_final) > 0) {
  lessThanMDL_final = lessThanMDL_final[, c(
    'lab_accession',
    'sta_seq',
    'chemparameter',
    'collect_date',
    'value',
    'MDL',
    'uom')]
}
```
```{r tables-lessThanMDL, fig.cap='Table 4. Table of values less than mdl', fig.align="center", echo=FALSE}
datatable(lessThanMDL_final, rownames = FALSE, options = list(pageLength = 5))
```


## Total vs Sum of Parts
The following is a table of values where the total value is compared to the sum of its parts. A record is considered flagged and put into this table if the sum of the components is more than the value marked as a total. Each record is organized by sharing a station sequence ID.

This script sums up the individual values of ammonia, nitrate, and nitrite to compare them to the entered value of total nitrogen. If the sum calculated by R is more than the value of total nitrogen, a flag is raised to mark the discrepancy. The script also compares the values of orthophosphate and total phosphorus, and raises another flag in the same situation. These comparisons are grouped by lab accession.

```{r sumOfParts, include=FALSE}
#create temporary sum table
sumOfParts <- table_sumOfParts

#create chem dataframe of specific chemical parameters
sumOfParts <- sumOfParts[sumOfParts$chemparameter == 'Ammonia' |
                sumOfParts$chemparameter == 'Nitrate' |
                sumOfParts$chemparameter == 'Nitrite' |
                sumOfParts$chemparameter == 'Ortho Phosphate' |
                sumOfParts$chemparameter == 'Total Nitrogen' |
                sumOfParts$chemparameter == 'Total Phosphorus', ]

#create list of unique station sequences
stations <- c(unique(sumOfParts$sta_seq))

#reset df_total
df_total = NULL

#convert null values to mdl
sumOfParts$value[is.na(sumOfParts$value)] <- sumOfParts$MDL

# loop through station sequences
for (n in 1:length(stations)) {
  #create temp dataframe of just this station sequence
  temp_sum <- sumOfParts[which(sumOfParts$sta_seq == stations[n]), ]
  
  #check that temp row has all needed chemical parameters
  if ('Ammonia' %in% temp_sum$chemparameter &
        'Nitrate' %in% temp_sum$chemparameter &
        'Nitrite' %in% temp_sum$chemparameter &
        'Ortho Phosphate' %in% temp_sum$chemparameter &
        'Total Nitrogen' %in% temp_sum$chemparameter &
        'Total Phosphorus' %in% temp_sum$chemparameter &
        'Ammonia' %in% temp_sum$chemparameter) {
    
    #aggregate sum by chemparameter
    chem_agg <- aggregate(temp_sum$value, by=list(Category=temp_sum$chemparameter), FUN=sum)
    
    #assign values
    ammonia <- chem_agg[1, 2]
    nitrate <- chem_agg[2, 2]
    nitrite <- chem_agg[3, 2]
    ortho   <- chem_agg[4, 2]
    tNitro  <- chem_agg[5, 2]
    tPhos   <- chem_agg[6, 2]
    
    #perform calculation
    sumDissolved <- ammonia + nitrate + nitrite
    
    chem_row = data.frame(
      'lab_accession' = temp_sum$lab_accession,
      'sta_seq' = temp_sum$sta_seq,
      ammonia,
      nitrate,
      nitrite,
      'Sum of Dissolved' = sumDissolved,
      tNitro,
      'Ortho Phosphate' = ortho,
      'Total Phosphorus' = tPhos)

    #add row to total dataframe
    df_total <- rbind(df_total, chem_row)
  }
}

#only keep records where sum of dissolved is more than total nitrogen
tempNitro <- df_total[which(df_total$Sum.of.Dissolved > df_table$tNitro), ]
#only keep records where sum of ortho phosphate is more than total phosphorus
tempPhos <- df_total[which(df_total$Ortho.Phosphate > df_table$tPhos), ]

#add to total dataframe
df_total <- rbind(df_total, tempNitro)
df_total <- rbind(df_total, tempPhos)
#wipe duplicates
df_total <- unique(df_total)

```
```{r tables-df_sumOfParts, fig.cap='Table 5. Table of total vs sum of parts', fig.align="center", echo=FALSE}
datatable(df_total, rownames = FALSE, options = list(pageLength = 5))
```


## Values Above 95th Percentile by Basin
The following is a table of chemical parameter values above the 95th percentile, by major basin.

```{r above95percentile, include=FALSE}
#Join in R syntax to create a table with sites and chemicals
table_basin <- merge(x=table_chem, y=sitebasin_base, by="sta_seq", all.x=TRUE)
#also combine with mdl
table_basin <- merge(x=table_basin, y=table_mdl, by='chemparameter', all.x=TRUE)

#declare complete df
allChem <- NULL

#create table of mbasn to major
basin <- unique(table_basin[, c('mbasn', 'major')])
#sort by mbasn
basin <- basin[order(basin$mbasn), ]
#convert to vector
basin_list <- c(basin$major)

for (i in 1:length(v_chems)) {
  #decide chemChoice
  chemChoice <- v_chems[i]
  
  #create temp data frame of chosen parameter
  df_chem <- table_basin[which(table_basin$chemparameter == chemChoice), ]
  df_total <- NULL #reset df_total at runtime
  
  #set mdl for chemparameter
  #inner_table_mdl <- table_mdl[table_mdl$chemparameter == chemChoice & table_mdl$MDL, ]
  ##update null values to mdl
  #df_chem$value[is.na(df_chem$value)] <- inner_table_mdl$MDL
  
  #loop through and organize by basins
  for (n in 1:length(basin_list)) {
    #create temp dataframe of current basin
    df_basin <- df_chem[df_chem$major == basin_list[n], ]
    
    if (nrow(df_basin) > 0) { #check if df_basin has any values
      #cbind basin name to table
      df_basin <- cbind(df_basin, basin_list[n])
      
      #find 95th percentile
      Q95 <- quantile(df_basin$value, .95, na.rm = TRUE)
      #cbind 95th percentile to the table
      df_basin <- cbind(df_basin, Q95)
      
      #keep only records where values above 95th percentile
      df_basin <- df_basin[df_basin$value > Q95, ]
      
      #add to to total dataframe
      df_total <- rbind(df_total, df_basin)
    }
  }
  #combine total of chem to all_chem table
  allChem <- rbind(allChem, df_total)
}

if (!is.null(nrow(allChem)) == TRUE && nrow(allChem) > 0) {
  #shorten allChem to specific fields
  test = allChem[, c('lab_accession',
                      'sta_seq',
                      'chemparameter',
                      'collect_date',
                      'name' = c(allChem$name.x),
                      'value',
                      'Q95',
                      'uom',
                      'major')]
  #omit null rows
  allChem <- na.omit(test)
}
```
```{r tables-above95th, fig.cap='Table 6. Table of values above 95th percentiles', fig.align="center", echo=FALSE}
datatable(allChem, rownames = FALSE, options = list(pageLength = 10))
```


## Values Above 95th Percentile by IC Group
The following is a table of chemical parameter values above the 95th percentile, organized by custom impervious cover (IC) metrics.

An important note is that this code portion relies on sites_Joined.csv.
The sites_Joined.csv table is generated from a join between Sites_HydroID and HSI_StreamCatch inside QGIS.

```{r above95percentileIC, include=FALSE}
#join t_chem and t_sites by sta_seq
total <- merge(table_chem, t_sites, by='sta_seq')
#also combine with mdl
total <- merge(total, table_mdl, by='chemparameter', all.x=TRUE)

#declare complete df
allChem <- NULL

#create table of custom IC Metric parameter values
customIC <- data.frame(
  'low' = c(0, 2, 10),
  'high' = c(2, 10, Inf)
)


for (i in 1:length(v_chems)) {
  #decide chemChoice
  chemChoice <- v_chems[i]
  
  #create temp data frame of chosen parameter
  df_chem <- total[total$chemparameter == chemChoice, ]
  df_total <- NULL #reset df_total at runtime
  
  ##update null values to mdl
  df_chem$value[is.na(df_chem$value)] <- df_chem$MDL
  
  #loop through ICMetric of 1 to 3, essentially grouping
  for (n in 1:3) {
    #create temp dataframe by custom ICMetric values
    df_IC <- df_chem[df_chem$IC_Avg > customIC$low[n] &
                       df_chem$IC_Avg < customIC$high[n], ]
    
    #check if there any rows in df_IC
    if (nrow(df_IC) > 0) {
      #!! IMPORTANT !!
      #change ICMetric to new categorization
      df_IC$ICMetric = n
      
      #find 95th percentile
      Q95 <- quantile(df_IC$value, .95, na.rm = TRUE)
      #add percentile as column
      df_IC <- cbind(df_IC, Q95)
      
      #keep only records where values above 95th percentile
      df_IC <- df_IC[df_IC$value > Q95, ]
      
      #add to to total dataframe
      df_total <- rbind(df_total, df_IC)
    }
  }
  #combine total of chem to all_chem table
  allChem <- rbind(allChem, df_total)
}

if (!is.null(nrow(allChem)) == TRUE && nrow(allChem) > 0) {
  #shorten allChem to specific fields
  test = allChem[, c('lab_accession',
                      'sta_seq',
                      'chemparameter',
                      'collect_date',
                      'name' = c(allChem$name.x),
                      'value',
                      'Q95',
                      'uom',
                      'ICMetric')]
  #omit null rows
  allChem <- na.omit(test)
}
```
```{r tables-above95thIC, fig.cap='Table 7. Table of values above 95th percentiles by IC group', fig.align="center", echo=FALSE}
datatable(allChem, rownames = FALSE, options = list(pageLength = 10))
```